---
title: "01-data_summary.Rmd"
author: "Claudia Romeo"
date: "2024-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Invasion hologenomics - Data Summary


## Load libraries

```{r load_libraries, include=FALSE}
library(R.utils)
library(knitr)
library(tidyverse)
library(dplyr)
library(devtools)
library(mapproj)

```

## Load data

```{r load_datasets, include=TRUE}

load("data/squirrels_data.Rdata")

```

## Sample summary

Summary of sampled individuals and analysed faecal samples.

```{r sample_description, message=FALSE}

#number of samples by species
sample_metadata %>%
  group_by(species,area_type) %>%
  summarise(n_samples = length(sample)) %>%
  kable()

#number of distinct squirrels
n_distinct(sample_metadata$animal)

#number of squirrels by species and type of area

sample_metadata %>%
  group_by(species,area_type) %>%
  summarise(distinct_squirrels = n_distinct(animal)) %>%
  kable()

#number of squirrels by species and season
sample_metadata %>%
  group_by(species,season) %>%
  summarise(distinct_squirrels = n_distinct(animal)) %>%
  kable()

#n of analysed faecal samples
ncol(read_counts)

```

Geographical location of sampled red squirrel (light blue) and grey squirrel (pink) populations in Italy.

```{r map, warning=FALSE}
#Summarise for generating map
options(dplyr.summarise.inform = FALSE)
sample_metadata_summary <- sample_metadata %>%
  #Group by geography and count samples
  select(sample, latitude, longitude, country, species) %>%
  group_by(latitude, longitude, species) %>%
  summarize(count = n()) %>%
  ungroup()

italy <- map_data("world", region="italy") %>%
  summarise(long = mean(long), lat = mean(lat))

#plotting on map
sample_metadata_summary %>%
  ggplot(.) +
  #render map
  geom_map(
    data=map_data("world", region="italy"),
    map = map_data("world", region="italy"),
    aes(long, lat, map_id=region),
    color = "white", fill = "#cccccc", linewidth = 0.2
  ) +
  #render points
  geom_point(
    aes(x=longitude,y=latitude, color=species),
    alpha=0.5, shape=16) +
  #add general plot layout
  theme_minimal() +
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        axis.title.y=element_blank()
  ) + coord_map("mercator")
```


```{r summary_table, include=FALSE}
#summary table
sample_metadata %>%
  select(sample,species,sample_type,country,latitude,longitude) %>%
  kable()

```

## Sequencing data summary

Total amount of sequencing data generated from the analysed samples.

```{r sequencing_summary}

#amount of discarded data (GB)
sum(round(((sample_metadata$metagenomic_bases+sample_metadata$host_bases)/
             (1-sample_metadata$bases_lost_fastp_percent))-
            (sample_metadata$metagenomic_bases+sample_metadata$host_bases)))/1000000000

#amount of host data (GB)
sum(sample_metadata$host_bases)/1000000000

#amount of metagenomic data (GB)
sum(sample_metadata$metagenomic_bases)/1000000000

#amount of estimated prokaryotic data (singleM)
sum(sample_metadata$metagenomic_bases * sample_metadata$singlem_fraction)/1000000000

```


Origin of DNA sequences obtained from each sample.

```{r sequence_origin}
sequence_fractions <- read_counts %>%
  pivot_longer(-genome, names_to = "sample", values_to = "value") %>%
  group_by(sample) %>%
  summarise(mags = sum(value)) %>%
  left_join(sample_metadata, by = join_by(sample == sample))  %>%
  select(sample,mags,metagenomic_bases,host_bases,bases_lost_fastp_percent) %>%
  mutate(mags_bases = mags*146) %>%
  mutate(lowqual_bases = ((metagenomic_bases+host_bases)/(1-bases_lost_fastp_percent))-(metagenomic_bases+host_bases)) %>%
  mutate(unmapped_bases = metagenomic_bases - mags_bases) %>%
  mutate(unmapped_bases = ifelse(unmapped_bases < 0, 0, unmapped_bases)) %>%
  select(sample,mags_bases,unmapped_bases,host_bases,lowqual_bases)

mags_bases_mean <- sequence_fractions %>%
  mutate(mags_bases = mags_bases / 1000000000) %>%
  select(mags_bases) %>%
  pull() %>%
  mean()

sequence_fractions %>%
  pivot_longer(!sample, names_to = "fraction", values_to = "value") %>%
  mutate(value = value / 1000000000) %>%
  mutate(fraction = factor(fraction, levels = c("lowqual_bases","host_bases","unmapped_bases","mags_bases"))) %>%
  ggplot(., aes(x = sample, y = value, fill=fraction)) +
  geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(values=c("#CCCCCC","#178a94","#ee8080","#d03161")) +
  geom_hline(yintercept = mags_bases_mean, linetype = "dashed", color = "black") +
  labs(x = "Samples", y = "Amount of data (GB)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),legend.position = "bottom")

```

## MAGs summary

```{r mags_summary}

#number of MAGs
nrow(read_counts)

#number of MAGs without species-level annotation (i.e., "new species")
genome_metadata %>%
  filter(species == "s__") %>%
  nrow()

#number of phylums
genome_metadata %>%
  select(phylum) %>%
  unique() %>%
  pull() %>%
  length()

```


