# Diversity analyses

```{r load_datasets04, include=FALSE, include=TRUE, message=FALSE, warning=FALSE}

rm(list=ls()) #clear environment

load("data/squirrels_data.Rdata")
singlem <- read.csv("data/singlem.csv",sep=";",header=T)
options(contrasts = c('contr.treatment','contr.poly'))

```

## Data preparation

```{r data_preparation04, message=FALSE}

#Change genome names column to row names
genome_counts <- genome_counts %>%
  column_to_rownames(var="genome")

#Get list of present MAGs
present_MAGs <- genome_counts %>%
  filter(rowSums(.[, -1]) != 0) %>%
  rownames()

#Remove samples with all zeros (no data after filtering)
genome_counts_filt <- genome_counts %>%
  select_if(~!all(. == 0))

#Align distillr annotations with present MAGs and remove all-zero and all-one traits
present_MAGs <- present_MAGs[present_MAGs %in% rownames(genome_gifts)]

genome_gifts_filt <- genome_gifts[present_MAGs,] %>%
  select_if(~!all(. == 0)) %>%  #remove all-zero modules
  select_if(~!all(. == 1)) #remove all-one modules

#Align tree with present MAGs
tree_filt <- keep.tip(genome_tree,present_MAGs)

#Filter count table to only contain present MAGs after gifts filtering
genome_counts_filt <- genome_counts[present_MAGs,]

#Calculate sequence fractions for each samples
sequence_fractions <- read_counts %>%
  pivot_longer(-genome, names_to = "sample", values_to = "value") %>%
  group_by(sample) %>%
  dplyr::summarize(mags = sum(value)) %>%
  left_join(sample_metadata, by = join_by(sample == sample))  %>%
  select(sample,mags,metagenomic_bases,host_bases,bases_lost_fastp_percent) %>%
  mutate(mags_bases = mags*146) %>%
  mutate(lowqual_bases = ((metagenomic_bases+host_bases)/(1-bases_lost_fastp_percent))-(metagenomic_bases+host_bases)) %>%
  mutate(unmapped_bases = metagenomic_bases - mags_bases) %>%
  mutate(unmapped_bases = ifelse(unmapped_bases < 0, 0, unmapped_bases)) %>%
  select(sample,mags_bases,unmapped_bases,host_bases,lowqual_bases)


```

## Alpha Diversity 

```{r alpha_diversity, cache=TRUE}
#ALPHA DIVERSITY
q0n <- hilldiv2::hilldiv(genome_counts,q=0) %>% c()
q1n <- hilldiv2::hilldiv(genome_counts,q=1) %>% c()
q1p <- hilldiv2::hilldiv(genome_counts,q=1,tree=genome_tree) %>% c()
dist <- hilldiv2::traits2dist(genome_gifts_filt, method="gower")
q1f <- hilldiv2::hilldiv(genome_counts_filt,q=1,dist=dist) %>% c()

```

```{r alpha_summary, warning=FALSE}

# Merge all metrics
alpha <- cbind(sample=colnames(genome_counts),richness=q0n,neutral=round(q1n,3),phylo=round(q1p,3),func=round(q1f,3)) %>%
  as.data.frame()
columns <- c("richness","neutral","phylo","func", "mapped","total")

# Add amount of sequencing data to the table
alpha <- alpha %>%
  left_join(sequence_fractions, by = join_by(sample == sample)) %>% #add sequencing depth information
  mutate(mapped=round(mags_bases/1000000000,3)) %>% #modify depth to million reads
  mutate(total=round((mags_bases+unmapped_bases+host_bases+lowqual_bases)/1000000000,3)) %>%
  select(sample,richness,neutral,phylo,func,mapped,total) %>%
  mutate(across(-1, as.numeric))

# Join sample metadata
alpha_div <- alpha %>%
  pivot_longer(-c(sample), names_to = "data", values_to = "value") %>%
  mutate(metric = factor(data, levels = columns)) %>%
  left_join(sample_metadata, by='sample')

 alpha_div %>%
  select(sample, species, metric, value) %>%
  mutate(
    species = factor(species), # Convert species to factor if necessary
    sample = factor(sample, levels = unique(sample[order(species)])) # Reorder sample by species
  ) %>%
  ggplot(aes(x = value, y = sample, fill = species)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = squirrel_colors) +
  facet_wrap(~ metric, scales = "free_x", ncol = 6) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    panel.spacing = unit(0, "lines"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_blank(),
    legend.position = "none"
  )
  

 
# #table
# kable(alpha_div)

```

### Alpha diversity comparisons

**Alpha diversity by species** 

```{r alpha_sp_plots, fig.dim = c(8, 10)}


neutral.sp <- alpha_div %>%
            filter(metric=="neutral")%>%
            ggboxplot(., x = "species", y = "value", color = "species", fill="white", add="jitter", alpha=0.5) +
                  scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
                  #stat_compare_means() +
                  theme_classic() +
                  labs(y = "Neutral Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())

phylo.sp <- alpha_div %>%
            filter(metric=="phylo")%>%
            ggboxplot(., x = "species", y = "value", color = "species", fill="white", add="jitter", alpha=0.5) +
                  scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
                  #stat_compare_means() +
                  theme_classic() +
                  labs(y = "Phylogenetic Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())

func.sp <- alpha_div %>%
            filter(metric=="func")%>%
            ggboxplot(., x = "species", y = "value", color = "species", fill="white", add="jitter") +
                  scale_color_manual(values=squirrel_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
                  #stat_compare_means() +
                  theme_classic() +
                  labs(y = "Functional Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())

ggarrange(neutral.sp, phylo.sp, func.sp, #+ rremove("x.text"), 
          ncol = 1, nrow = 3)
```

**Alpha diversity by species and sex**

```{r alpha_sex_plots, fig.dim = c(8, 10)}



neutral.sex <- alpha_div %>%
            filter(metric=="neutral") %>%
            ggboxplot(., x = "species", y = "value", color = "sex", fill="white", add="jitter") +
                  scale_color_manual(values=sex_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
                  #stat_compare_means() +
                  theme_classic() +
                  labs(y = "Neutral Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())

phylo.sex <- alpha_div %>%
            filter(metric=="phylo") %>%
            ggboxplot(., x = "species", y = "value", color = "sex", fill="white", add="jitter") +
                  scale_color_manual(values=sex_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
                  #stat_compare_means() +
                  theme_classic() +
                  labs(y = "Phylogenetic Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())


func.sex <- alpha_div %>%
            filter(metric=="func") %>%
            ggboxplot(., x = "species", y = "value", color = "sex", fill="white", add="jitter") +
                  scale_color_manual(values=sex_colors) +
                  scale_fill_manual(values=paste0(squirrel_colors)) +
                  #stat_compare_means() +
                  theme_classic() +
                  labs(y = "Functional Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())


sex.legend <- get_legend(neutral.sex)


ggarrange(neutral.sex, phylo.sex, func.sex, #+ rremove("x.text"), 
          legend.grob = sex.legend, legend="right", common.legend = TRUE,
          #labels = c("A", "B", "C"),
          ncol = 1, nrow = 3)

neutral.sex.test <- alpha_div %>%
            filter(metric=="neutral")
wilcox.test(as.numeric(value) ~ sex, data=neutral.sex.test, exact = FALSE) %>%
            print()

phylo.sex.test <- alpha_div %>%
            filter(metric=="phylo")
wilcox.test(as.numeric(value) ~ sex, data=phylo.sex.test, exact = FALSE) %>%
            print()

func.sex.test <- alpha_div %>%
            filter(metric=="func")
wilcox.test(as.numeric(value) ~ sex, data=phylo.sex.test, exact = FALSE) %>%
            print()


```


**Alpha diversity by species and urbanisation**

```{r alpha_urb_plots, fig.dim = c(8, 10)}

sample_metadata$area_type <-factor(sample_metadata$area_type, levels = c("rural", "suburban", "urban"))

#neutral alpha by species*area_type
neutral.urb <- alpha_div %>%
            filter(metric=="neutral") %>%
            ggboxplot(., x = "species", y = "value", color = "area_type", fill="white", add="jitter") +
            scale_color_manual(values=area_colors) +
            scale_fill_manual(values=paste0(area_colors)) +
            stat_compare_means() +
            theme_classic() +
            labs(y = "Neutral Hill numbers") +
            theme(
              legend.position = "none",
              axis.title.x = element_blank()) +
            guides(color=guide_legend(title="Urbanisation"), fill="none")

#phylogenetic alpha by species*area_type
phylo.urb <- alpha_div %>%
            filter(metric=="phylo") %>%
            ggboxplot(., x = "species", y = "value", color = "area_type", fill="white", add="jitter") +
                  scale_color_manual(values=area_colors) +
                  scale_fill_manual(values=paste0(area_colors)) +
                  stat_compare_means() +
                  theme_classic() +
                  labs(y = "Phylogenetic Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())

#functional (distillr-based) alpha by species*area_type
func.urb <- alpha_div %>%
            filter(metric=="func") %>%
            ggboxplot(., x = "species", y = "value", color = "area_type", fill="white", add="jitter") +
                  scale_color_manual(values=area_colors) +
                  scale_fill_manual(values=paste0(area_colors)) +
                  stat_compare_means() +
                  theme_classic() +
                  labs(y = "Functional Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())


urb.legend <- get_legend(neutral.urb)


ggarrange(neutral.urb, phylo.urb, func.urb, #+ rremove("x.text"), 
          legend.grob = urb.legend, legend="right", common.legend = TRUE,
          #labels = c("A", "B", "C"),
          ncol = 1, nrow = 3)
```


**Alpha diversity by species and season**

```{r alpha_season_plots, fig.dim = c(8, 10)}

sample_metadata$season <-factor(sample_metadata$season, levels = c("spring-summer", "autumn", "winter"))

#neutral alpha by species*season
neutral.seas <- alpha_div %>%
            filter(metric=="neutral") %>%
            ggboxplot(., x = "species", y = "value", color = "season", fill="white", add="jitter") +
            scale_color_manual(values=season_colors) +
            scale_fill_manual(values=paste0(season_colors)) +
            stat_compare_means() +
            theme_classic() +
            labs(y = "Neutral Hill numbers") +
            theme(
              legend.position = "none",
              axis.title.x = element_blank()) +
            guides(color=guide_legend(title="Season"), fill="none")

#phylogenetic alpha by species*season
phylo.seas <- alpha_div %>%
            filter(metric=="phylo") %>%
            ggboxplot(., x = "species", y = "value", color = "season", fill="white", add="jitter") +
                  scale_color_manual(values=season_colors) +
                  scale_fill_manual(values=paste0(season_colors)) +
                  stat_compare_means() +
                  theme_classic() +
                  labs(y = "Phylogenetic Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())

#functional (distillr-based) alpha by species*season
func.seas <- alpha_div %>%
            filter(metric=="func") %>%
            ggboxplot(., x = "species", y = "value", color = "season", fill="white", add="jitter") +
                  scale_color_manual(values=season_colors) +
                  scale_fill_manual(values=paste0(season_colors)) +
                  stat_compare_means() +
                  theme_classic() +
                  labs(y = "Functional Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())


seas.legend <- get_legend(neutral.seas)


ggarrange(neutral.seas, phylo.seas, func.seas, #+ rremove("x.text"), 
          legend.grob = seas.legend, legend="right", common.legend = TRUE,
          #labels = c("A", "B", "C"),
          ncol = 1, nrow = 3)

```


**Alpha diversity by species and development**

```{r alpha_dev_plots, fig.dim = c(8, 10)}

#neutral alpha by species*development
neutral.dev <- alpha_div %>%
            filter(metric=="neutral") %>%
            ggboxplot(., x = "species", y = "value", color = "development", fill="white", add="jitter") +
            stat_compare_means() +
            theme_classic() +
            labs(y = "Neutral Hill numbers") +
            theme(
              legend.position = "none",
              axis.title.x = element_blank()) +
            guides(color=guide_legend(title="Development"), fill="none")

#phylogenetic alpha by species*development
phylo.dev <- alpha_div %>%
            filter(metric=="phylo") %>%
            ggboxplot(., x = "species", y = "value", color = "development", fill="white", add="jitter") +
                  stat_compare_means() +
                  theme_classic() +
                  labs(y = "Phylogenetic Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())

#functional (distillr-based) alpha by species*development
func.dev <- alpha_div %>%
            filter(metric=="func") %>%
            ggboxplot(., x = "species", y = "value", color = "development", fill="white", add="jitter") +
                  stat_compare_means() +
                  theme_classic() +
                  labs(y = "Functional Hill numbers") +
                  theme(
                      legend.position = "none",
                      axis.title.x = element_blank())


dev.legend <- get_legend(neutral.dev)


ggarrange(neutral.dev, phylo.dev, func.dev, #+ rremove("x.text"), 
          legend.grob = dev.legend, legend="right", common.legend = TRUE,
          #labels = c("A", "B", "C"),
          ncol = 1, nrow = 3)


```



### Alpha diversity bayesian models

**Data preparation for bayesian models**

```{r data_prep_models}

diversity.data <- alpha %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  mutate(season=factor(season, levels = c("spring-summer", "autumn", "winter"))) %>%
  right_join(singlem, by = join_by(sample == sample)) %>%
  group_by(species) %>%
  mutate(index500_st = scale(index500, center=T, scale=T)[,1]) %>%
  ungroup() %>%
  filter(sample!="EHI02263") %>% #remove outlier
  filter(development=="Adult") #remove juveniles, nursing and pregnant females

#check whether a low domain-adjusted mapping rate (DAMR) is associated with low diversity estimates 
ggplot(diversity.data, aes(x=est_mapp, y=neutral)) +
  geom_point(size=3, alpha=0.5, color="#6c9ebc") +
  labs(x = "DAMR (mapping rate to MAG catalogue/singleM microbial fraction estimate)", y = "Neutral diversity (effective number of MAGs)") +
  theme_classic() +
  theme(legend.position="none")

diversity.data <- diversity.data %>%
  filter(mags_singlem > 0.8) #remove 5 samples with low DAMR


#str(diversity.data)

```


**Bayesian models - neutral alpha**

```{r neutral_brm, message=FALSE, warning=FALSE, results='hide', cache=TRUE}

set.seed(123)
neutral.brm <-brm(neutral ~ species + index500 + season + 
            + species:index500 + species:season 
            + (1|animal) + (1|sampling_site), 
            data=diversity.data,
            family=gaussian(),
            chains = 3,
            iter = 3000, 
            warmup = 1000)

# library(report)
# report(neutral.brm, verbose = FALSE)

# #assessing model fit
# plot(neutral.brm) 

# #posterior predictive check
# pp_check(neutral.brm)

```

```{r neutral_brm_res, message=FALSE, warning=FALSE}


describe_posterior(neutral.brm)

plot(p_direction(neutral.brm), stack=FALSE)

#conditional effects
neutral_eff <- (conditional_effects(neutral.brm, effects = "species:index500"))
neutral_df <- as.data.frame(neutral_eff$'species:index500')

neutral_df %>% 
  ggplot(., aes(x=index500, y=estimate__, color=species, fill=species)) + 
  geom_line(size=0.9) +
  geom_ribbon(aes(ymin=lower__, ymax=upper__), alpha=0.35, linetype=0) +
  scale_color_manual(values= squirrel_colors) + 
  scale_fill_manual(values=squirrel_colors) +
  theme_minimal() +
  labs(x = "Urbanization index", y = "Neutral diversity estimate")


```



**Bayesian models - phylogenetic alpha**

```{r phylo_brm, message=FALSE, warning=FALSE, results='hide'}

set.seed(123)
phylo.brm <-brm(phylo ~ species + index500 + season + 
            + species:index500 + species:season 
            + (1|animal) + (1|sampling_site),
            data=diversity.data,
            family=gaussian(),
            chains = 3,
            iter = 3000, 
            warmup = 1000)


# #assessing model fit
# plot(phylo.brm) #pairs(phylo.brm)

# #posterior predictive check
# pp_check(phylo.brm)


```

```{r phylo_brm_res, message=FALSE, warning=FALSE}

describe_posterior(phylo.brm)

plot(p_direction(phylo.brm), stack=FALSE)

#conditional effects
phylo_eff <- (conditional_effects(phylo.brm, effects = "species:index500"))
phylo_df <- as.data.frame(phylo_eff$'species:index500')

phylo_df %>% 
  ggplot(., aes(x=index500, y=estimate__, color=species, fill=species)) + 
  geom_line(size=0.9) +
  geom_ribbon(aes(ymin=lower__, ymax=upper__), alpha=0.35, linetype=0) +
  scale_color_manual(values= squirrel_colors) + 
  scale_fill_manual(values=squirrel_colors) +
  theme_minimal() +
  labs(x = "Urbanization index", y = "Phylogenetic diversity estimate")


```


**Bayesian models - functional alpha**


```{r func_brm, message=FALSE, warning=FALSE, results='hide'}

set.seed(123)
func.brm <-brm(func ~ species + index500 + season + 
            + species:index500 + species:season 
            + (1|animal),
            data=diversity.data,
            family=gaussian(),
            chains = 3,
            iter = 3000, 
            warmup = 1000,
            control = list(adapt_delta = 0.9))


# #assessing model fit
# plot(func.brm) 
# 
# #posterior predictive check
# pp_check(func.brm)


```

```{r func_brm_res, message=FALSE, warning=FALSE}

describe_posterior(func.brm)

plot(p_direction(func.brm), stack=FALSE)

# names(func.brm$fit)
# func.brm$fit <- func.brm$fit %>% 
#   recode("b_season1" = "b_seasonautumn",
#                             "b_season2" = "b_seasonwinter")
# 
# 
# # Obtain the p_direction data
# pd_data <- p_direction(func.brm)
# 
# unique(pd_data$Parameter)
# 
# parnames(func.brm)
# tidybayes::get_variables(func.brm)
# 
# func.brm2 <- rename_pars(func.brm)
# 
# # Manually set the labels
# # Assuming "season1" corresponds to "seasonautumn" and "season2" to "seasonwinter"
# pd_data$Parameter <- recode(pd_data$Parameter,
#                             "b_season1" = "b_seasonautumn",
#                             "b_season2" = "b_seasonwinter")
# 
# pd_data$Parameter <- gsub("b_season1", "b_seasonautumn", pd_data$Parameter)
# pd_data$Parameter <- gsub("b_season2", "b_seasonwinter", pd_data$Parameter)
# 
# plot(pd_data)

#conditional effects
func_eff <- (conditional_effects(func.brm, effects = "species:season"))
func_df <- as.data.frame(func_eff$'species:season')

func_df %>% 
  ggplot(., aes(x=species, y=estimate__, group=season, color=season, fill=season)) + 
  geom_point(position=position_dodge(width=0.4), size=4) +
  geom_errorbar(aes(ymin=lower__, ymax=upper__, group=season), position=position_dodge(width=0.4), size=0.6, width=0.4) +
  scale_color_manual(values= season_colors) + 
  scale_fill_manual(values=season_colors) +
  theme_classic() +
  labs(x = "", y = "Functional diversity estimate")

```


## Beta diversity

```{r beta_diversity04, message=FALSE, eval=TRUE, cache=TRUE}

#neutral beta div ALL
adult <- sample_metadata %>% 
  filter(development=="Adult") %>%
  pull(sample)
genome_counts_adult <- genome_counts %>%
  select(all_of(adult))
beta_q1n <-hilldiv2::hillpair(genome_counts_adult, q=1, metric="S")
# beta_q1p <-hilldiv2::hillpair(genome_counts_adult,q=1,tree=genome_tree, metric="S")
# beta_q1f <-hilldiv2::hillpair(genome_counts_adult,q=1,dist=dist)

#neutral beta div RED
red <- sample_metadata %>% 
  filter(development=="Adult") %>%
  filter(species=="Sciurus vulgaris") %>%
  pull(sample)
genome_counts_red <- genome_counts %>%
  select(all_of(red))
beta_red <-hilldiv2::hillpair(genome_counts_red, q=1, metric="S")

#neutral beta div GREY
grey <- sample_metadata %>% 
  filter(development=="Adult") %>%
  filter(species=="Sciurus carolinensis") %>%
  pull(sample)
genome_counts_grey <- genome_counts %>%
  select(all_of(grey))
beta_grey <-hilldiv2::hillpair(genome_counts_grey, q=1, metric="S")

```

### Both species

```{r permanova_beta, eval=TRUE}

#neutral beta diversity PERMANOVA
sample_metadata_adonis <- sample_metadata %>%
  filter(sample %in% labels(beta_q1n)) %>%
  arrange(sample) %>%
  #mutate(location=paste0(round(longitude,2),"_",round(latitude,2))) %>%
  select(animal,sample,species,index500,season,sex) %>%
  select_if(~ length(unique(.)) > 1) %>% #remove columns with all-identical values
  column_to_rownames(var = "sample") %>%
  as.data.frame()


dispersion <- betadisper(beta_q1n, group=sample_metadata_adonis[labels(beta_q1n),]$species)
# plot(dispersion, main="PCoA")
boxplot(dispersion$distances ~ dispersion$group, main="Distance to centroids")
permutest(dispersion)

#adjust permutation design to account for repeated measures in subsequent PERMANOVA
hr <- with(sample_metadata_adonis, how(nperm = 999, blocks = animal)) 

adonis2(formula=beta_q1n ~ species + index500 + season + sex, 
        data=sample_metadata_adonis[labels(beta_q1n),], permutations=hr, na.action = na.omit) %>%
  as.matrix() %>%
  print()

pairwise.adonis(beta_q1n,sample_metadata_adonis[labels(beta_q1n),]$season, perm=999)

```

```{r NMDS_all, eval=TRUE, results='hide'}

#NMDS
set.seed(123)
beta_q1n_nmds <- beta_q1n %>%
  metaMDS(.,trymax = 500, k=2, verbosity=FALSE) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  filter(sample !="EHI00420")  #remove outlier

```

```{r NMDS_all_plot, fig.width=10, fig.height=8}

beta_q1n_nmds %>% group_by(species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(., aes(x = NMDS1, y = NMDS2)) + 
  scale_colour_manual(values = squirrel_colors) +
  scale_fill_manual(values = squirrel_colors) +
  geom_point(size = 3, alpha=0.8, aes(color = species)) + 
  stat_ellipse(data=beta_q1n_nmds, aes(x = NMDS1, y = NMDS2, group = species, fill=species, color=species),geom="polygon",level=0.95,alpha=0, size=0.8) + 
  geom_segment(aes(x=x_cen, y=y_cen, xend=NMDS1, yend=NMDS2, color=species), alpha=0.3) +
  labs(x = "NMDS1", y = "NMDS2") +
  theme_classic() +
  theme(axis.text = element_text(colour = "black", size = 12),
        axis.title = element_text(size = 14, colour = "black"), 
        legend.text = element_text(size = 11, colour ="black"), 
        legend.position = "right", axis.title.y = element_text(size = 14), 
        legend.title = element_text(size = 14, colour = "black"),
        legend.key=element_blank(),
        panel.background = element_blank(), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5))


```


### Red squirrel

```{r permanova_beta_red, eval=TRUE}

metadata_adonis_red <- sample_metadata %>%
  filter(sample %in% labels(beta_red)) %>%
  arrange(sample) %>%
  #mutate(location=paste0(round(longitude,2),"_",round(latitude,2))) %>%
  select(animal,sample,species,index500,season,sex) %>%
  select_if(~ length(unique(.)) > 1) %>% #remove columns with all-identical values
  column_to_rownames(var = "sample") %>%
  as.data.frame()


#adjust permutation design to account for repeated measures in subsequent PERMANOVA
hr_red <- with(metadata_adonis_red, how(nperm = 999, blocks = animal)) 

adonis2(formula=beta_red ~ season + index500 + sex, 
        data=metadata_adonis_red[labels(beta_red),], permutations=hr_red, na.action = na.omit) %>%
  as.matrix() %>%
  print()


dispersion_red <- betadisper(beta_red, group=metadata_adonis_red[labels(beta_red),]$season)
# plot(dispersion_red, main="PCoA")
boxplot(dispersion_red$distances ~ dispersion_red$group, main="Distance to centroids")
permutest(dispersion_red)

pairwise.adonis(beta_red,metadata_adonis_red[labels(beta_red),]$season, perm=999)


```


```{r NMDS_red, eval=TRUE, results='hide'}


#NMDS
set.seed(123)
beta_red_nmds <- beta_red %>%
  metaMDS(.,trymax = 500, k=2, verbosity=FALSE) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  filter(sample !="EHI00420")  #remove outliers

#stressplot(beta_q1n_nmds)


```

```{r NMDS_red_plot, fig.width=10}

# Extract vector and factors from NMDS ordination
env <- beta_red_nmds %>% select(index500, season)
nmds <- beta_red_nmds %>% select(NMDS1, NMDS2)
en <- envfit(nmds, env, permutations = 999, na.rm = TRUE)
en

# ordisurf(x = nmds,
# y = env$index500,
# bubble = 4)


# Make dataframe with vector to add to ggplot
vec.df <- as.data.frame(en$vectors$arrows*sqrt(en$vectors$r))
vec.df$variables <- rownames(vec.df)


beta_red_nmds %>% group_by(season) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(., aes(x = NMDS1, y = NMDS2)) + 
  scale_colour_manual(values = season_colors) +
  scale_fill_manual(values = season_colors) +
  stat_ellipse(aes(x = NMDS1, y = NMDS2, group = season, color=season, fill=season), geom="polygon", alpha=0, size=0.8, level=0.95) + 
  geom_point(size = 3, alpha=1, aes(color = season)) + 
  geom_segment(data = vec.df,
               aes(x = 0, xend = NMDS1*0.1, y = 0, yend = NMDS2*0.1),
               arrow = arrow(length = unit(0.25, "cm")),
               inherit.aes = FALSE, size =0.6, alpha = 1, colour = "grey10") + 
  #geom_segment(aes(x=x_cen, y=y_cen, xend=NMDS1, yend=NMDS2, color=season), alpha=0.3) +
  geom_text(data = vec.df,
            aes(x = NMDS1*0.12, y=NMDS2*0.12, label = variables),
            size=4.5) +
  labs(x = "NMDS1", y = "NMDS2") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black", size = 12),
        axis.title = element_text(size = 14, colour = "black"), 
        legend.text = element_text(size = 11, colour ="black"), 
        legend.position = "right", 
        legend.key=element_blank(),
        legend.title = element_text(size = 14, colour = "black"),
        panel.background = element_blank(), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5))

```


### Grey squirrel

```{r permanova_beta_grey, eval=TRUE}

#PERMANOVA
metadata_adonis_grey <- sample_metadata %>%
  filter(sample %in% labels(beta_grey)) %>%
  arrange(sample) %>%
  #mutate(location=paste0(round(longitude,2),"_",round(latitude,2))) %>%
  select(animal,sample,species,index500,season,sex) %>%
  select_if(~ length(unique(.)) > 1) %>% #remove columns with all-identical values
  column_to_rownames(var = "sample") %>%
  as.data.frame()


#adjust permutation design to account for repeated measures in subsequent PERMANOVA
hr_grey <- with(metadata_adonis_grey, how(nperm = 999, blocks = animal)) 

adonis2(formula=beta_grey ~ season + index500 + sex, 
        data=metadata_adonis_grey[labels(beta_grey),], permutations=hr_grey, na.action = na.omit) %>%
  as.matrix() %>%
  print()

dispersion_grey <- betadisper(beta_grey, group=metadata_adonis_grey[labels(beta_grey),]$season)
# plot(dispersion_grey, main="PCoA")
boxplot(dispersion_grey$distances ~ dispersion_grey$group, main="Distance to centroids")
permutest(dispersion_grey)

pairwise.adonis(beta_grey,metadata_adonis_grey[labels(beta_grey),]$season, perm=999)


```


```{r NMDS_grey, eval=TRUE, results='hide'}

#NMDS
set.seed(123)
beta_grey_nmds <- beta_grey %>%
  metaMDS(.,trymax = 900, k=2, verbosity=FALSE) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  filter(sample !="EHI00420")  #remove outliers

#stressplot(beta_q1n_nmds)

```

```{r NMDS_grey_plot, fig.width=10}


# Extract vector and factors from NMDS ordination
env <- beta_grey_nmds %>% select(index500, season)
nmds <- beta_grey_nmds %>% select(NMDS1, NMDS2)

en <- envfit(nmds, env, permutations = 999, na.rm = TRUE)
en

# ordisurf(x = nmds,
# y = env$index500,
# bubble = 4)

# Make dataframe with vector to add to ggplot
vec.df <- as.data.frame(en$vectors$arrows*sqrt(en$vectors$r))
vec.df$variables <- rownames(vec.df)


beta_grey_nmds %>% group_by(season) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(., aes(x = NMDS1, y = NMDS2)) + 
  scale_colour_manual(values = season_colors) +
  scale_fill_manual(values = season_colors) +
  stat_ellipse(aes(x = NMDS1, y = NMDS2, group = season, color=season, fill=season), geom="polygon", alpha=0, size=0.8, level=0.95) + 
  geom_point(size = 3, alpha=1, aes(color = season)) + 
  geom_segment(data = vec.df,
               aes(x = 0, xend = NMDS1*0.1, y = 0, yend = NMDS2*0.1),
               arrow = arrow(length = unit(0.25, "cm")),
               inherit.aes = FALSE, size =0.6, alpha = 1, colour = "grey10") + 
  #geom_segment(aes(x=x_cen, y=y_cen, xend=NMDS1, yend=NMDS2, color=season), alpha=0.3) +
  geom_text(data = vec.df,
            aes(x = NMDS1*0.51, y=NMDS2*0.51, label = variables),
            size=4.5) +
  labs(x = "NMDS1", y = "NMDS2") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  theme_classic() +
  theme(axis.text = element_text(colour = "black", size = 12),
        axis.title = element_text(size = 14, colour = "black"), 
        legend.text = element_text(size = 11, colour ="black"), 
        legend.position = "right", 
        legend.key=element_blank(),
        legend.title = element_text(size = 14, colour = "black"),
        panel.background = element_blank(), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5))


```

