---
title: "05-Diversity models"
author: "Claudia Romeo"
date: "2023-12-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Load libraries

```{r load_libraries, include=TRUE, message=FALSE, warning=FALSE}

library(lme4)
library(emmeans)
library(effects)
library(car)
library(lmerTest)
library(ggResidpanel)
library(tidyverse)
library(dplyr)
library(gtsummary)
library(broom.mixed)

options(contrasts = c('contr.sum','contr.poly'))
```

## Load data

IMPORTANT: to generate alpha and beta diversity estimates, script 03-diversity_analysis.Rmd must be run first 

```{r load_datasets, include=TRUE}

load("data/squirrels_data.Rdata")

alpha_div <- read.csv("data/alpha_div.csv",sep=";",header=T)
singlem <- read.csv("data/singlem.csv",sep=";",header=T)


```

## Data preparation

```{r data_prep, include=TRUE}

diversity.data <- alpha_div %>%
  # mutate(log.neut = log(phylo))   %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  mutate(season=factor(season, levels = c("spring-summer", "autumn", "winter"))) %>%
  #left_join(singlem,sample_metadata, by = join_by(sample == sample)) %>%
  right_join(singlem, by = join_by(sample == sample)) %>%
  group_by(species) %>%
  mutate(index500_st = scale(index500, center=T, scale=T)[,1]) %>%
  ungroup() %>%
  filter(sample!="EHI02263") 


diversity.data %>% group_by(development) %>% 
 mutate(neutral=as.numeric(neutral)) %>% mean(neutral, na.rm=TRUE)

diversity.data %>% 
  group_by(development) %>% 
  summarize(median.n = median(neutral),
            median.f = median (func2)) 

plot(diversity.data$mags_singlem, diversity.data$neutral)  


#str(diversity.data)

```

```{r var_distrib, eval=FALSE}

# check y distributions
plot(diversity.data$neutral)
hist(diversity.data$neutral, breaks=30)
d <- density(diversity.data$neutral)
plot(d)

plot(diversity.data$phylo)
hist(diversity.data$phylo, breaks=30)
d <- density(diversity.data$phylo)
plot(d)

plot(diversity.data$func2)
hist(diversity.data$func2, breaks=30)
d <- density(diversity.data$func2)
plot(d)

```



## LMM - neutral alpha

### Neutral model - area_type

```{r neutral_model, eval=FALSE}

neutral.model<-lmer(neutral ~ species + area_type + season 
            + species:area_type + species:season
            + (1|animal) + (1|macroarea), 
            data=diversity.data, na.action=na.omit, 
            control=lmerControl(optimizer="bobyqa", 
                                optCtrl=list(maxfun=2e5)))
summary(neutral.model)
Anova(neutral.model, test.statistic='F', type=3)

```


```{r neutral_diagnostics, include=FALSE, eval=FALSE}

#check_model(neutral.model) #not giving the full panel in rmd
resid_panel(neutral.model, smoother = TRUE, qqbands = TRUE)

```

```{r neutral_posthoc, eval=FALSE}
# post hoc testing
lsmeans(neutral.model, 'species')
lsmeans(neutral.model, pairwise ~ species, adjust='holm')

lsmeans(neutral.model, 'area_type')
lsmeans(neutral.model, pairwise ~ area_type, adjust='holm')
```

```{r neutral_plots, eval=FALSE}

neutral.model.eff <- allEffects(neutral.model) 
plot(neutral.model.eff)

plot(effect('species', neutral.model))
plot(effect('area_type', neutral.model))

```


### Neutral model - index500

```{r neutral_model2}

neutral.model2<-lmer(neutral ~ species + index500 + season + development
            + species:index500 + species:season 
            + (1|animal),
            data=diversity.data, na.action=na.omit,
            control=lmerControl(optimizer="bobyqa",
                                optCtrl=list(maxfun=2e5)))
summary(neutral.model2)
Anova(neutral.model2, test.statistic='F', type=3)

#check_model(neutral.model) #not giving the full panel in rmd
resid_panel(neutral.model2, smoother = TRUE, qqbands = TRUE)

```
```{r neutral_plots2}

neutral.model.eff2 <- allEffects(neutral.model2) 
plot(neutral.model.eff2)

# plot(effect('species', neutral.model2))
# plot(effect('builtup_st', neutral.model2))

```

## LMM - phylogenetic alpha

### Phylo model - area_type

```{r phylo_model, eval=FALSE}

phylo.model<-lmer(phylo ~ species + area_type + season 
                    + species:area_type + species:season 
                    + (1|animal) + (1|macroarea), 
                    data=diversity.data, na.action=na.omit, 
                    control=lmerControl(optimizer="bobyqa", 
                                        optCtrl=list(maxfun=2e5)))

summary(phylo.model)
Anova(phylo.model, test.statistic='F', type=3)

```

```{r phylo_diagnostics, eval=FALSE}

resid_panel(phylo.model, smoother = TRUE, qqbands = TRUE)

```

```{r phylo_posthoc, eval=FALSE}

lsmeans(phylo.model, 'species')
lsmeans(phylo.model, pairwise ~ species, adjust='holm')

lsmeans(phylo.model, 'area_type')
lsmeans(phylo.model, pairwise ~ area_type, adjust='holm')

```

```{r phylo_plots, eval=FALSE}

phylo.model.eff <- allEffects(phylo.model) 
plot(phylo.model.eff)

plot(effect('species', phylo.model))
plot(effect('area_type', phylo.model))

```

### Phylo model - built-up fraction

```{r phylo_model2}

phylo.model2<-lmer(phylo ~ species + index500 + season
                    + species:index500 + species:season 
                    + (1|animal),
                    data=diversity.data, na.action=na.omit,
                    control=lmerControl(optimizer="bobyqa",
                                        optCtrl=list(maxfun=2e5)))

summary(phylo.model2)
Anova(phylo.model2, test.statistic='F', type=3)

resid_panel(phylo.model2, smoother = TRUE, qqbands = TRUE)

```


```{r phylo_plots2}

phylo.model.eff2 <- allEffects(phylo.model2) 
plot(phylo.model.eff2)
# 
# plot(effect('species', phylo.model2))
# plot(effect('area_type', phylo.model2))

```

## LMM - functional (distillr-based) alpha

### Func model - area_type

```{r func2_model, eval=FALSE}

func2.model<-lmer(func2 ~ species + area_type + season 
                  + species:area_type + species:season +
                  + (1|animal) , 
                  data=diversity.data, na.action=na.omit, 
                  control=lmerControl(optimizer="bobyqa", 
                                      optCtrl=list(maxfun=2e5)))

summary(func2.model)
Anova(func2.model, test.statistic='F',  type=3)

```

```{r func2_diagnostics, eval=FALSE}

resid_panel(func2.model, smoother = TRUE, qqbands = TRUE)

```

```{r func2_posthocs, eval=FALSE}

lsm_area <- emmeans::lsmeans(func2.model, specs= ~ species:area_type)
Sc.nat <- c(1, 0, 0, 0, 0, 0)
Sc.semi <- c(0, 0, 1, 0, 0, 0)
Sc.urb <- c(0, 0, 0, 0, 1, 0)
Sv.nat <- c(0, 1, 0, 0, 0, 0)
Sv.semi <- c(0, 0, 0, 1, 0, 0)
Sv.urb <- c(0, 0, 0, 0, 0, 1)
emmeans::contrast(lsm_area, method = list("Natural: Sc vs Sv" = Sc.nat - Sv.nat,
                             "Seminatural: Sc vs Sv" = Sc.semi - Sv.semi,
                             "Urban: Sc vs Sv" = Sc.urb - Sv.urb,
                             "Sc: natural vs semi" = Sc.nat - Sc.semi,
                             "Sc: natural vs urban" = Sc.nat - Sc.urb,
                             "Sc: semi vs urban" = Sc.semi - Sc.urb,
                             "Sv: natural vs semi" = Sv.nat - Sv.semi,
                             "Sv: natural vs urban" = Sv.nat - Sv.urb,
                             "Sv: semi vs urban" = Sv.semi - Sv.urb), adjust='holm')

lsm_seas <- emmeans::lsmeans(func2.model, specs= ~ species:season)
Sc.ss <- c(1, 0, 0, 0, 0, 0)
Sc.au <- c(0, 0, 1, 0, 0, 0)
Sc.wi <- c(0, 0, 0, 0, 1, 0)
Sv.ss <- c(0, 1, 0, 0, 0, 0)
Sv.au <- c(0, 0, 0, 1, 0, 0)
Sv.wi <- c(0, 0, 0, 0, 0, 1)
emmeans::contrast(lsm_seas, method = list("Spring-summer: Sc vs Sv" = Sc.ss - Sv.ss,
                             "Autumn: Sc vs Sv" = Sc.au - Sv.au,
                             "Winter: Sc vs Sv" = Sc.wi - Sv.wi,
                             "Sc: spring-sum vs aut" = Sc.ss - Sc.au,
                             "Sc: spring-sum vs win" = Sc.ss - Sc.wi,
                             "Sc: aut vs win" = Sc.au - Sc.wi,
                             "Sv: spring-sum vs aut" = Sv.ss - Sv.au,
                             "Sv: spring-sum vs win" = Sv.ss - Sv.wi,
                             "Sv: aut vs win" = Sv.au - Sv.wi), adjust='holm')

```

```{r func2_plots, eval=FALSE}

func2.model.eff <- allEffects(func2.model) 
plot(func2.model.eff)

```


### Func model - built-up fraction


```{r func2_model2}

func2.model2<-lmer(func2 ~ species + index500 + season + development
                  + species:index500 + species:season 
                  + (1|animal),
                  data=diversity.data, na.action=na.omit,
                  control=lmerControl(optimizer="bobyqa",
                                      optCtrl=list(maxfun=2e5))) 
  

summary(func2.model2)
Anova(func2.model2, test.statistic='F', type=3)

resid_panel(func2.model2, smoother = TRUE, qqbands = TRUE)

# func2.model2 %>% tbl_regression() %>%
#   add_global_p()

```


```{r func2_plots2}

func2.model.eff2 <- allEffects(func2.model2)
plot(func2.model.eff2)

```