# Joint Species Distribution Modelling - output analysis

```{r load_data07}

rm(list=ls()) #clear environment
load("data/squirrels_data.Rdata")
options(contrasts = c('contr.sum','contr.poly'))

# Select desired support threshold
support_threshold=0.9
negsupport_threshold=1-support_threshold
```

```{r hmsc_model_list}

# Select modelchain of interest
load("hmsc/fit_model.3a_250_10.Rdata")

levels.1a <- c("species","index500","season","logseqdepth","Random: animal")
levels.1b <- c("species","index500","season","logseqdepth","Random: animal", "Random: sampling_site")
levels.2a <- c("species","index500","season","species:index500","logseqdepth","Random: animal")
levels.2b <- c("species","index500","season","species:index500","logseqdepth","Random: animal", "Random: sampling_site")
levels.3a <- c("species","index500","season","species:season","logseqdepth","Random: animal")
levels.3b <- c("species","index500","season","species:season","logseqdepth","Random: animal", "Random: sampling_site")
levels.4a <- c("species","index500","season","species:index500","species:season","logseqdepth","Random: animal")
levels.4b <- c("species","index500","season","logseqdepth","species:index500","species:season","Random: animal", "Random: sampling_site")
```

## Variance partitioning

```{r hmsc_varpart, warning=FALSE, comments="", message=FALSE, cache=TRUE}
# Compute variance partitioning
varpart=computeVariancePartitioning(m)

varpart$vals %>%
   as.data.frame() %>%
   rownames_to_column(var="variable") %>%
   pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
   mutate(variable=factor(variable, levels=levels.3a)) %>%
   group_by(variable) %>%
   summarise(mean=mean(value)*100,sd=sd(value)*100) %>%
   tt()
```

```{r hmsc_varpart_plot, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10)}
# Basal tree
varpart_tree <- genome_tree %>%
        keep.tip(., tip=m$spNames) 

# Varpart table
varpart_table <- varpart$vals %>%
   as.data.frame() %>%
   rownames_to_column(var="variable") %>%
   pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
   mutate(variable=factor(variable, levels=rev(levels.3a))) %>%
   mutate(genome=factor(genome, levels=rev(varpart_tree$tip.label)))

# Phylums
phylums <- genome_metadata %>%
    filter(genome %in% varpart_tree$tip.label) %>%
    arrange(match(genome, varpart_tree$tip.label)) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome") %>%
    select(phylum)

# Basal ggtree
varpart_tree <- varpart_tree %>%
        force.ultrametric(.,method="extend") %>%
        ggtree(., size = 0.3)

# Add phylum colors next to the tree tips
varpart_tree <- gheatmap(varpart_tree, phylums, offset=-0.2, width=0.1, colnames=FALSE) +
   scale_fill_manual(values=custom_colors)+
      labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
varpart_tree <- varpart_tree + new_scale_fill()

# Add variance stacked barplot
vertical_tree <-  varpart_tree +
        scale_fill_manual(values=c("#83bb90","#cccccc","#ed8a45","#b2b530","#be3e2b","#12738f","#f6de9c")) + #"#122f3d"
        geom_fruit(
             data=varpart_table,
             geom=geom_bar,
             mapping = aes(x=value, y=genome, fill=variable, group=variable),
             pwidth = 2,
             offset = 0.05,
             width= 1,
             orientation="y",
             stat="identity")+
      labs(fill="Variable")

vertical_tree
```

## Posterior estimates

```{r hmsc_postestimates, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10)}
# Select desired support threshold
support=0.9
negsupport=1-support

# Basal tree
postestimates_tree <- genome_tree %>%
        keep.tip(., tip=m$spNames) 

#plotBeta(hM=m, post=getPostEstimate(hM=m, parName="Beta"), param = "Support", plotTree = TRUE, covNamesNumbers=c(1,0))

# Posterior estimate table
post_beta <- getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
    mutate(genome=factor(genome, levels=rev(postestimates_tree$tip.label))) %>%
    mutate(value = case_when(
          value >= support ~ "Positive",
          value <= negsupport ~ "Negative",
          TRUE ~ "Neutral")) %>%
    mutate(value=factor(value, levels=c("Positive","Neutral","Negative"))) %>%
    pivot_wider(names_from = variable, values_from = value) %>%
    rename(intercept=2,
            Sc=3,
            index500=4,
            autumn=5,
            winter=6,
            logseqdepth=7,
            "Sc:autumn"=8,
            "Sc:winter"=9) %>%
   column_to_rownames(var="genome") %>%
  select(-intercept) 
  
#Phylums
phylums <- genome_metadata %>%
    filter(genome %in% postestimates_tree$tip.label) %>%
    arrange(match(genome, varpart_tree$tip.label)) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome") %>%
    select(phylum)


# Basal ggtree
postestimates_tree <- postestimates_tree %>%
        force.ultrametric(.,method="extend") %>%
        ggtree(., size = 0.3)
        

#Add phylum colors next to the tree tips
postestimates_tree <- gheatmap(postestimates_tree, phylums, offset=-0.2, width=0.1, colnames=FALSE) +
      scale_fill_manual(values=custom_colors)+
      labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
postestimates_tree <- postestimates_tree + new_scale_fill()

# Add posterior significant heatmap

postestimates_tree <- gheatmap(postestimates_tree, post_beta, offset=0, width=0.5, colnames=TRUE, colnames_position="top",colnames_angle=90, colnames_offset_y=1, hjust=0) +
        scale_fill_manual(values=c("#be3e2b","#f4f4f4","#b2b530"))+
        labs(fill="Trend") 

postestimates_tree +
        vexpand(.30, 1)  # expand top 

```

## Model fit

```{r model_fit_R2, message=FALSE, warning=FALSE}

#NB: where does cv come from?? ask Ostaizka

# MFCV <- evaluateModelFit(hM=m, predY=cv)
# 
# mean(MFCV$R2, na.rm = TRUE)
# 
# genome_fit <- tibble(genome=m$spNames, r2 = MFCV[[2]])
# 
# genome_counts_filt %>% 
# mutate_if(is.numeric, ~ . / sum(.)) %>% 
# left_join(genome_fit, by="genome") %>% 
# filter(r2>0.5) %>% 
# select(-c(genome,r2)) %>% 
# colSums() %>%  
# hist()
# 
# var_pred_table <- tibble(mag=m$spNames,
#        pred=MFCV$R2,
#        var_pred=MFCV$R2 * varpart$vals[1,],
#        support=getPostEstimate(hM=m, parName="Beta")$support %>% .[2,],
#        estimate=getPostEstimate(hM=m, parName="Beta")$mean %>% .[2,]) %>%
#   mutate(enrichment=ifelse(support>=support_threshold,"Feral","Neutral")) %>%
#   mutate(enrichment=ifelse(support<=negsupport_threshold,"Domestic",enrichment))
# 
# var_pred_table %>%
#   ggplot(aes(x=estimate,y=var_pred, color=enrichment))+
#     geom_point()+
#     scale_color_manual(values=c("#bd70ae","#949293","#ebe8e8"))+
#     geom_hline(yintercept=0.005, linetype="dashed")+
#     theme_minimal()
# 
# predictive_mags <- var_pred_table %>% 
#   filter(var_pred>=0.005) %>% 
#   pull(mag)
```


## Correlations among bacteria

```{r hmsc_correlations, warning=FALSE, comments="", message=FALSE, fig.dim=c(8,8)}
#Compute the residual correlation matrix
OmegaCor = computeAssociations(m)

# Reference tree (for sorting genomes)
genome_tree_subset <- genome_tree %>%
        keep.tip(., tip=m$spNames) 


#Co-occurrence matrix at the animal level
supportLevel = 0.95
toPlot = ((OmegaCor[[1]]$support>supportLevel)
          + (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean

matrix <- toPlot %>% 
      as.data.frame() %>%
      rownames_to_column(var="genome1") %>%
      pivot_longer(!genome1, names_to = "genome2", values_to = "cor") %>%
      mutate(genome1= factor(genome1, levels=genome_tree_subset$tip.label)) %>%
      mutate(genome2= factor(genome2, levels=genome_tree_subset$tip.label)) %>%
      ggplot(aes(x = genome1, y = genome2, fill = cor)) +
            geom_tile() + 
            scale_fill_gradient2(low = "#be3e2b",
                       mid = "#f4f4f4",
                       high = "#b2b530")+
            theme_void() +
            theme(legend.position = "none") 

corr.legend <- get_legend(matrix, position="right") 
corr.legend <- as_ggplot(corr.legend)

vtree <- genome_tree_subset %>%
  force.ultrametric(.,method="extend") %>%
  ggtree(., expand=1.5) +
  hexpand(0.5)

#Add phylum colors next to the tree tips
vtree <- gheatmap(vtree, phylums, offset=-0.2, width=0.6, colnames=FALSE) +
      scale_fill_manual(values=custom_colors) +
      theme(legend.position = 'none')

vtreeD <- genome_tree_subset %>%
  force.ultrametric(.,method="extend") %>%
  ggtree(., expand=1.5, layout="dendrogram") 

#Add phylum colors next to the tree tips
vtreeD <- gheatmap(vtreeD, phylums, offset=-0.2, width=0.8, colnames=FALSE) +
      scale_fill_manual(values=custom_colors) +
      theme(legend.position = 'none')


#create composite figure
grid.arrange(grobs = list(vtreeD,matrix,vtree,corr.legend),
             layout_matrix = rbind(c(5,5,1,1,1,1,1,1,1,1,1,5),
                                   c(5,5,1,1,1,1,1,1,1,1,1,5),
                                   c(5,5,1,1,1,1,1,1,1,1,1,5),
                                   c(3,3,2,2,2,2,2,2,2,2,2,5),
                                   c(3,3,2,2,2,2,2,2,2,2,2,5),
                                   c(3,3,2,2,2,2,2,2,2,2,2,5),
                                   c(3,3,2,2,2,2,2,2,2,2,2,4),
                                   c(3,3,2,2,2,2,2,2,2,2,2,4),
                                   c(3,3,2,2,2,2,2,2,2,2,2,4),
                                   c(3,3,2,2,2,2,2,2,2,2,2,4),
                                   c(3,3,2,2,2,2,2,2,2,2,2,5),
                                   c(3,3,2,2,2,2,2,2,2,2,2,5),
                                   c(3,3,2,2,2,2,2,2,2,2,2,5),
                                   c(3,3,2,2,2,2,2,2,2,2,2,5)))

```

## Association with host species

```{r}

estimate <- getPostEstimate(hM=m, parName="Beta")$mean %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    filter(variable=="speciesSciurus carolinensis") %>%
    pivot_longer(!variable, names_to = "genome", values_to = "mean") %>%
    dplyr::select(genome,mean)

support <- getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    filter(variable=="speciesSciurus carolinensis") %>%
    pivot_longer(!variable, names_to = "genome", values_to = "support") %>%
    dplyr::select(genome,support)

inner_join(estimate,support,by=join_by(genome==genome)) %>%
    mutate(significance=ifelse(support >= 0.9 | support <= 0.1,1,0)) %>%
    mutate(support=ifelse(mean<0,1-support,support)) %>%
    left_join(genome_metadata, by = join_by(genome == genome)) %>%
    mutate(phylum = ifelse(support > 0.9, phylum, NA)) %>%
    ggplot(aes(x=mean,y=support,color=phylum))+
      geom_point(alpha=0.7, shape=16, size=3)+
      scale_color_manual(values = alpha(custom_colors, 0.7)) +
      geom_vline(xintercept = 0) +
      xlim(c(-2,2)) +
      labs(x = "Beta regression coefficient", y = "Posterior probability") +
      theme_minimal()
```



### MAGs associated with **S. carolinensis**

```{r}

getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
    mutate(trend = case_when(
          value >= support_threshold ~ "Positive",
          value <= negsupport_threshold ~ "Negative",
          TRUE ~ "Neutral")) %>%
    filter(variable=="speciesSciurus carolinensis", trend=="Positive") %>%
    arrange(-value) %>%
    left_join(genome_metadata,by=join_by(genome==genome)) %>%
    dplyr::select(genome,phylum,class,order,family,species,value) %>%
    arrange(phylum, class, family, species)%>%
  paged_table()

```
### MAGs associated with **S. vulgaris**

```{r}

getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
    mutate(trend = case_when(
          value >= support_threshold ~ "Positive",
          value <= negsupport_threshold ~ "Negative",
          TRUE ~ "Neutral")) %>%
    filter(variable=="speciesSciurus carolinensis", trend=="Negative") %>%
    arrange(-value) %>%
    left_join(genome_metadata,by=join_by(genome==genome)) %>%
    dplyr::select(genome,phylum,class,order,family,species,value) %>%
    arrange(phylum, class, family, species)%>%
  paged_table()

```

## Microbiome composition predictions

### Predicted composition by species

```{r hmsc_pred_species, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10)}

# Overall species prediction
pred_species <- constructGradient(m, 
                      focalVariable = "species", 
                      non.focalVariables = list(logseqdepth=list(1),index500=list(1), season=list(1)), 
                      ngrid=gradientlength) %>%
             predict(m, Gradient = ., expected = TRUE) %>%
             as.data.frame() %>%
             mutate(species=c("Sciurus vulgaris","Sciurus carolinensis")) %>%
             pivot_longer(!species, names_to = "genome", values_to = "value") %>%
             mutate(genome = sub("(.*\\..*\\.)[^.]+.*", "\\1", genome)) %>% #remove iteration suffix
             mutate(genome =  sub("\\.$", "", genome))

# levels(m$XData$species)
# levels(m$XData$season)

pred_species %>%
      pivot_wider(names_from = species, values_from = value) %>%
      unnest(c(`Sciurus carolinensis`, `Sciurus vulgaris`)) %>%
      mutate(diff.grey_red = `Sciurus carolinensis` - `Sciurus vulgaris`) %>%
      select(genome, diff.grey_red) %>%
      left_join(genome_metadata, by=join_by(genome==genome)) %>%
      mutate(genome= factor(genome, levels=genome_tree_subset$tip.label)) %>%
      ggplot(., aes(y=genome, x=diff.grey_red, fill=phylum, color=phylum)) +
          scale_color_manual(values=custom_colors)+
          geom_vline(xintercept = 0)+
          scale_fill_manual(values=alpha(custom_colors,0.3))+
          geom_boxplot(outlier.shape = NA) +
          theme_classic() +
          theme(axis.text.y = element_blank())

```

Support values of the host species variable per genome are interpreted as enrichment for either red (S. vulgaris) or grey squirrels (S. carolinensis)

```{r hmsc_sp_enrichment, warning=FALSE, comments="", message=FALSE}
species_enrichment <- getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
    mutate(enrichment = case_when(
          value >= support_threshold ~ "Sciurus carolinensis",
          value <= negsupport_threshold ~ "Sciurus vulgaris",
          TRUE ~ "Neutral")) %>%
    filter(variable=="speciesSciurus carolinensis") %>%
    select(genome, enrichment)
```

The most likely microbiome compositions of red (S. vulgaris) and grey squirrels (S. carolinensis) as predicted by the hmsc model

```{r sp_prediction, warning=FALSE, comments="", message=FALSE}
gradient = c("Sciurus vulgaris","Sciurus carolinensis")
gradientlength = length(gradient)

pred_sp <- constructGradient(m, 
                      focalVariable = "species", 
                      non.focalVariables = 1) %>%
            predict(m, Gradient = ., expected = TRUE) %>%
            do.call(rbind,.) %>%
            as.data.frame() %>%
            mutate(species=rep(gradient,1000)) %>%
            pivot_longer(!species,names_to = "genome", values_to = "value")
```

Log-abundance differences between Sciurus vulgaris and S. carolinensis for genera found enriched in either species.

```{r sp_plot, comment="", message=FALSE, warning=FALSE, fig.dim=c(8,10), fig.fullwidth=TRUE}

genome_metadata_clean <- genome_metadata %>%  
    mutate(family = coalesce(family, paste('Unclassified', order)),
           genus = coalesce(genus, 
                              if_else(grepl('^Unclassified', family),
                                      family, paste('Unclassified', family))))

pred_sp %>%
  left_join(species_enrichment,by="genome") %>% 
      filter(enrichment != "Neutral") %>% 
      pivot_wider(names_from = species, values_from = value) %>%
      unnest(c(`Sciurus vulgaris`, `Sciurus carolinensis`)) %>%
      mutate(diff_Sv.Sc = `Sciurus vulgaris`-`Sciurus carolinensis`) %>%
      select(genome, diff_Sv.Sc) %>%
      left_join(genome_metadata_clean, by="genome") %>%
      mutate(genome= factor(genome, levels=genome_tree$tip.label)) %>%
      ggplot(., aes(y=forcats::fct_reorder(genus,diff_Sv.Sc), x=diff_Sv.Sc, fill=phylum, color=phylum)) +
            geom_boxplot(outlier.shape = NA) +
            scale_color_manual(values=custom_colors)+
            scale_fill_manual(values=alpha(custom_colors,0.4))+
          geom_vline(xintercept = 0)+
          theme_classic()+
      labs(x = "Grey-red abundance difference", y = "Genera")


```


### Predicted composition by urbanization

```{r hmsc_pred_urb, warning=FALSE, message=FALSE}

gradient = seq(0,1, by=0.1)
gradientlength = length(gradient)

pred_urban <- constructGradient(m,
                      focalVariable = "index500",
                      non.focalVariables = list(logseqdepth=list(1),species=list(1), season=list(1)),
                      ngrid=gradientlength) %>%
                      predict(m, Gradient = ., expected = TRUE) %>%
                      do.call(rbind,.) %>%
                      as.data.frame() %>%
                      mutate(index500=rep(gradient,1000)) %>%
                      pivot_longer(-c(index500), names_to = "genome", values_to = "value")

getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
    mutate(trend = case_when(
          value >= support_threshold ~ "Positive",
          value <= negsupport_threshold ~ "Negative",
          TRUE ~ "Neutral")) %>%
    filter(variable=="index500") %>%
    select(genome,trend) %>%
    left_join(pred_urban, by=join_by(genome==genome)) %>%
    group_by(genome, trend, index500) %>%
    summarize(value = mean(value, na.rm = TRUE)) %>%
    left_join(genome_metadata, by=join_by(genome == genome)) %>%
    ggplot(aes(x=index500, y=value, group=genome, color=phylum, linetype=trend)) +
        geom_line() +
        scale_linetype_manual(values=c("solid","dashed","solid")) +
        scale_color_manual(values=custom_colors) +
        facet_grid(fct_rev(trend) ~ phylum) +
        labs(y="Genome abundance (log)",x="Urbanization index") +
        theme(legend.position = "none") +
        theme_minimal() +
         theme(legend.position = "none",
               axis.text.x = element_text(angle = 45, hjust = 0.8,),
               axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
               strip.text.x = element_text(angle = 90, hjust = 0,),
               strip.text.y = element_text(face="bold"),
)
```

### Predicted composition by season

```{r hmsc_pred_season, warning=FALSE, comments="", message=FALSE, fig.dim=c(8, 10)}

# pred_season_Sc <- constructGradient(m,
#                       focalVariable = "season",
#                       non.focalVariables = list(logseqdepth=list(1),species=list(3, "Sciurus carolinensis"), index500=list(1)),
#                       ngrid=gradientlength) %>%
#              predict(m, Gradient = ., expected = TRUE) %>%
#              as.data.frame() %>%
#              mutate(season=c("spring-summer","autumn","winter")) %>%
#              pivot_longer(!season, names_to = "genome", values_to = "value")
# 
# getPostEstimate(hM=m, parName="Beta")$support %>%
#     as.data.frame() %>%
#     mutate(variable=m$covNames) %>%
#     pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
#     mutate(trend = case_when(
#           value >= support_threshold ~ "Positive",
#           value <= negsupport_threshold ~ "Negative",
#           TRUE ~ "Neutral")) %>%
#     filter(variable %in% c("seasonspring-summer", "seasonautumn", "seasonwinter")) %>%
#     select(genome,trend) %>%
#     left_join(pred_season_Sc, by=join_by(genome==genome)) %>%
#     group_by(genome, trend, season) %>%
#     summarize(value = mean(value, na.rm = TRUE)) %>%
#     left_join(genome_metadata, by=join_by(genome == genome)) %>%
#     ggplot(aes(x=season, y=value, group=genome, color=phylum, linetype=trend)) +
#         geom_boxplot(outlier.shape = NA)  +
#         scale_color_manual(values=custom_colors) +
#         facet_grid(fct_rev(trend) ~ phylum) +
#         labs(y="Genome abundance (log)",x="Season") +
#         theme(legend.position = "none") +
#         theme_minimal() +
#          theme(legend.position = "none",
#                axis.text.x = element_text(angle = 45, hjust = 0.8,),
#                axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
#                strip.text.x = element_text(angle = 90, hjust = 0,)
# )
# 
# pred_season %>%
#       mutate(genome= factor(genome, levels=genome_tree_subset$tip.label)) %>%
#       group_by(genome, season) %>%
#       summarize(value = mean(value, na.rm = TRUE)) %>%
#       left_join(genome_metadata, by=join_by(genome==genome)) %>%
#       filter(!is.na(phylum)) %>%
#       ggplot(., aes(y=value, x=genome, group=phylum, color=phylum, fill=phylum)) +
#       facet_grid( . ~ season) +
#       scale_color_manual(values=custom_colors)+
#       scale_fill_manual(values=alpha(custom_colors,0.3))+
#       geom_boxplot(outlier.shape = NA) +
#       theme_classic() +
#       theme(axis.text.y = element_blank(),
#             axis.text.x = element_blank())
    
```

## Microbiome functional predictions

```{r GIFTs_data_preparation2, echo = T, results = 'hide', warning=FALSE, message=FALSE}

tss <- function(abund){sweep(abund, 2, colSums(abund), FUN="/")} 

genome_counts <- genome_counts %>%
  column_to_rownames(var="genome")
genome_kegg <- genome_kegg %>%
  column_to_rownames(var="genome")
genome_gifts <- genome_gifts %>%
  column_to_rownames(var="genome")

#Get list of present MAGs
present_MAGs <- genome_counts %>%
  filter(rownames(.) %in% m$spNames) %>%
  rownames()

#Align distillr annotations with present MAGs and remove all-zero and all-one traits
present_MAGs <- present_MAGs[present_MAGs %in% rownames(genome_gifts)]

GIFTs_elements <- genome_gifts[present_MAGs,] %>%
  select_if(~!all(. == 0)) %>%  #remove all-zero modules
  select_if(~!all(. == 1)) #remove all-one modules

#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements,GIFT_db)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs and get overall metabolic capacity indices per MAG (at the domain level)
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db) %>% as.data.frame() %>%
  mutate(Overall=rowMeans(select(.,Biosynthesis,Structure,Degradation), na.rm=TRUE))

```


### Function-level predictions

#### Predicted functions by species

```{r function_pred_sp, message=FALSE, warning=FALSE}

community_func_sp <- pred_sp %>%
  filter(genome %in% present_MAGs) %>% #keep only predictive mags -> NB: I can't get predictive mags rn (see Model fit)
  group_by(species, genome) %>%
  mutate(row_id = row_number()) %>%
  pivot_wider(names_from = genome, values_from = value) %>%
  ungroup() %>%
  group_split(row_id) %>%
  as.list() %>%
  lapply(., FUN = function(x){x %>%
    dplyr::select(-row_id) %>%
    column_to_rownames(var = "species") %>%
    as.data.frame() %>%
    exp() %>%
    t() %>%
    tss() %>%
    to.community(GIFTs_functions,.,GIFT_db) %>% 
    as.data.frame() %>%
    rownames_to_column(var="species")
   })

function_pred_sp <- map_dfr(community_func_sp, function(df) {
  Sc_values <- df %>% filter(species == "Sciurus carolinensis") %>% select(-species)
  Sv_values <- df %>% filter(species == "Sciurus vulgaris") %>% select(-species)
  Sc_values - Sv_values
}) %>%
  mutate(iteration=c(1:1000)) %>% 
  pivot_longer(!iteration,names_to="trait",values_to="value") %>%
      group_by(trait) %>%
      summarise(mean=mean(value),
        p1 = quantile(value, probs = 0.1),
        p9 = quantile(value, probs = 0.9),
        positive_support = sum(value > 0)/1000,
        negative_support = sum(value < 0)/1000) %>%
      arrange(-positive_support)

```

```{r function_sp_tables, message=FALSE, warning=FALSE}

# Positively associated
function_pred_sp %>%
  filter(mean >0) %>%
  arrange(-positive_support) %>%
  filter(positive_support>=0.9) %>%
  paged_table()

# Negatively associated
function_pred_sp %>%
  filter(mean <0) %>%
  arrange(-negative_support) %>%
  filter(negative_support>=0.9) %>%
  paged_table()

```

```{r function_sp_plot, message=FALSE, warning=FALSE, fig.dim=c(10, 8)}

positive <- function_pred_sp %>%
  filter(mean >0) %>%
  arrange(mean) %>%
  filter(positive_support>=0.9) %>%
  dplyr::select(-negative_support) %>%
  rename(support=positive_support)

negative <- function_pred_sp %>%
  filter(mean <0) %>%
  arrange(mean) %>%
  filter(negative_support>=0.9) %>%
  dplyr::select(-positive_support) %>%
  rename(support=negative_support)


all_functions <- function_pred_sp %>%
  left_join(GIFT_db,by=join_by(trait==Code_function)) %>%
  mutate(trait=factor(trait)) %>%
  mutate(function_legend=str_c(trait," - ",Function)) %>%
  select(trait,mean,p1,p9,function_legend) %>% 
  unique()

gift_colors <- gift_colors %>% 
  mutate(legend=str_c(Code_function," - ",Function))  %>% 
  filter(legend %in% all_functions$function_legend)

all_functions %>%
  ggplot(aes(x=mean, y=fct_reorder(function_legend, mean), xmin=p1, xmax=p9, color=function_legend)) +
      geom_point() +
      geom_errorbar() +
      #xlim(c(-0.12,0.05)) +
      geom_vline(xintercept=0) +
      scale_color_manual(values = gift_colors$Color) +
      theme_minimal() +
      labs(x="Regression coefficient",y="Functional trait") +
      #guides(col = guide_legend(ncol = 1))
      theme(legend.position = "none")

```

#### Predicted functions by urbanization

```{r function_urb_pred, results="hide", eval=TRUE, message=FALSE, warning=FALSE, cache=TRUE}

community_func_urb <- pred_urban %>%
  group_by(index500, genome) %>%
  mutate(row_id = row_number()) %>%
  pivot_wider(names_from = genome, values_from = value) %>%
  ungroup() %>%
  group_split(row_id) %>%
  as.list() %>%
  lapply(., FUN = function(x){x %>%
    select(-row_id) %>%
    column_to_rownames(var = "index500") %>%
    as.data.frame() %>%
    exp() %>%
    t() %>%
    tss() %>%
    to.community(GIFTs_functions,.,GIFT_db) %>%
    as.data.frame() %>%
    rownames_to_column(var="index500")
   })


calculate_slope <- function(x) {
  lm_fit <- lm(unlist(x) ~ seq_along(unlist(x)))
  coef(lm_fit)[2]
}

function_pred_urb <- map_dfc(community_func_urb, function(mat) {
      mat %>%
        column_to_rownames(var = "index500") %>%
        t() %>%
        as.data.frame() %>%
        rowwise() %>%
        mutate(slope = calculate_slope(c_across(everything()))) %>%
        select(slope) }) %>%
      t() %>%
      as.data.frame() %>%
      set_names(colnames(community_func_urb[[1]])[-1]) %>%
      rownames_to_column(var="iteration") %>%
      pivot_longer(!iteration, names_to="trait",values_to="value") %>%
      group_by(trait) %>%
      summarise(mean=mean(value),
        p1 = quantile(value, probs = 0.1),
        p9 = quantile(value, probs = 0.9),
        positive_support = sum(value > 0)/1000,
        negative_support = sum(value < 0)/1000) %>%
      arrange(-positive_support)

```

```{r function_urb_tables, message=FALSE, warning=FALSE}

# Positively associated
function_pred_urb %>%
  filter(mean >0) %>%
  arrange(-positive_support) %>%
  filter(positive_support>=0.9) %>%
  paged_table()

#Negatively associated
function_pred_urb %>%
  filter(mean <0) %>%
  arrange(-negative_support) %>%
  filter(negative_support>=0.9) %>%
  paged_table()

```

```{r function_urb_plot, message=FALSE, warning=FALSE, fig.dim=c(10, 8)}
#Positively associated
positive <- function_pred_urb%>%
  filter(mean >0) %>%
  arrange(mean) %>%
  filter(positive_support>=0.9) %>%
  select(-negative_support) %>%
  rename(support=positive_support)

#Negatively associated
negative <- function_pred_urb %>%
  filter(mean <0) %>%
  arrange(mean) %>%
  filter(negative_support>=0.9) %>%
  select(-positive_support) %>%
  rename(support=negative_support)

all_functions <- function_pred_urb %>%
  left_join(GIFT_db,by=join_by(trait==Code_function)) %>%
  mutate(trait=factor(trait)) %>%
  mutate(function_legend=str_c(trait," - ",Function)) %>%
  select(trait,mean,p1,p9,function_legend) %>% 
  unique()

gift_colors <- read_tsv("data/gift_colors.tsv") %>% 
  mutate(legend=str_c(Code_function," - ",Function))  %>% 
  filter(legend %in% all_functions$function_legend)

all_functions %>%
  ggplot(aes(x=mean, y=fct_reorder(function_legend, mean), xmin=p1, xmax=p9, color=function_legend)) +
      geom_point() +
      geom_errorbar() +
      xlim(c(-0.017,0.017)) +
      geom_vline(xintercept=0) +
      scale_color_manual(values = gift_colors$Color) +
      theme_minimal() +
      labs(x="Regression coefficient",y="Functional trait") +
      guides(col = guide_legend(ncol = 1))

# all_functions <- bind_rows(positive,negative) %>%
#   left_join(GIFT_db,by=join_by(trait==Code_function)) %>%
#   mutate(trait=factor(trait,levels=c(rev(positive$trait),rev(negative$trait)))) %>%
#   mutate(Code_function=factor(Code_function)) %>%
#   mutate(function_legend=str_c(Code_function," - ",Function)) %>%
#   select(trait,mean,p1,p9,function_legend) %>% 
#   unique()
# 
# gift_colors <- gift_colors %>% 
#   mutate(legend=str_c(Code_function," - ",Function))  %>% 
#   filter(legend %in% all_elements$function_legend)
# 
# all_functions %>%
#   ggplot(aes(x=mean, y=fct_reorder(function_legend, mean), xmin=p1, xmax=p9, color=function_legend)) +
#       geom_point() +
#       geom_errorbar() +
#       #xlim(c(-0.05,0.05)) +
#       geom_vline(xintercept=0) +
#       scale_color_manual(values = gift_colors$Color) +
#       theme_minimal() +
#       labs(x="Regression coefficient",y="Functional trait")

community_func_urb %>%
    bind_rows() %>%
    pivot_longer(-index500, names_to = "trait", values_to = "value") %>%
    filter(trait %in% c(positive$trait, negative$trait)) %>%
    mutate(trait=factor(trait, levels=c(positive$trait, negative$trait))) %>%
    mutate(index500=as.numeric(index500)) %>%
    ggplot(aes(x=index500, y=value)) +
          geom_smooth(method = lm, formula = y ~ x, se = TRUE) +
          #geom_smooth(method = lm, formula = y ~ splines::bs(x, 3), se = TRUE) +
          facet_wrap(~trait, ncol=5, scales="free") +
          theme_minimal() +
          labs(x="Urbanization",y="Metabolic Capacity Index") 


```

#### Predicted elements by season

**S. carolinensis**

**S. vulgaris**


### Element-level predictions

#### Predicted elements by species

```{r element_sp_pred, results="hide", eval=TRUE, message=FALSE, warning=FALSE}

community_elem_sp <- pred_sp %>%
  group_by(species, genome) %>%
  mutate(row_id = row_number()) %>%
  pivot_wider(names_from = genome, values_from = value) %>%
  ungroup() %>%
  group_split(row_id) %>%
  as.list() %>%
  lapply(., FUN = function(x){x %>%
    select(-row_id) %>%
    column_to_rownames(var = "species") %>%
    as.data.frame() %>%
    exp() %>%
    t() %>%
    tss() %>%
    to.community(GIFTs_elements,.,GIFT_db) %>%
    as.data.frame() %>%
    rownames_to_column(var="species")
   })

element_pred_sp <- map_dfr(community_elem_sp, function(df) {
  Sc_values <- df %>% filter(species == "Sciurus carolinensis") %>% select(-species)
  Sv_values <- df %>% filter(species == "Sciurus vulgaris") %>% select(-species)
  Sc_values - Sv_values
}) %>%
  mutate(iteration=c(1:1000)) %>% 
  pivot_longer(!iteration,names_to="trait",values_to="value") %>%
      group_by(trait) %>%
      summarise(mean=mean(value),
        p1 = quantile(value, probs = 0.1),
        p9 = quantile(value, probs = 0.9),
        positive_support = sum(value > 0)/1000,
        negative_support = sum(value < 0)/1000) %>%
      arrange(-positive_support)

# Positively associated
element_pred_sp %>%
  filter(mean >0) %>%
  arrange(-positive_support) %>%
  filter(positive_support>=0.9) %>%
  paged_table()

# Negatively associated
element_pred_sp %>%
  filter(mean <0) %>%
  arrange(-negative_support) %>%
  filter(negative_support>=0.9) %>%
  paged_table()



```
```{r element_sp_plot, message=FALSE, warning=FALSE, fig.dim=c(10, 12)}

positive <- element_pred_sp %>%
  filter(mean >0) %>%
  arrange(mean) %>%
  filter(positive_support>=0.9) %>%
  dplyr::select(-negative_support) %>%
  rename(support=positive_support)

negative <- element_pred_sp %>%
  filter(mean <0) %>%
  arrange(mean) %>%
  filter(negative_support>=0.9) %>%
  dplyr::select(-positive_support) %>%
  rename(support=negative_support)

all_elements <- bind_rows(positive,negative) %>%
  left_join(GIFT_db,by=join_by(trait==Code_element)) %>%
  mutate(trait=factor(trait,levels=c(rev(positive$trait),rev(negative$trait)))) %>%
  mutate(Code_function=factor(Code_function)) %>%
  mutate(element_legend=str_c(trait," - ",Element)) %>%
  mutate(function_legend=str_c(Code_function," - ",Function)) %>%
  select(trait,mean,p1,p9,element_legend,function_legend) %>% 
  unique()

gift_colors <- read_tsv("data/gift_colors.tsv") %>% 
  mutate(legend=str_c(Code_function," - ",Function))  %>% 
  filter(legend %in% all_elements$function_legend)

all_elements %>%
  ggplot(aes(x=mean, y=fct_reorder(element_legend, mean), xmin=p1, xmax=p9, color=function_legend)) +
      geom_point() +
      geom_errorbar() +
      #xlim(c(-0.05,0.05)) +
      geom_vline(xintercept=0) +
      scale_color_manual(values = gift_colors$Color) +
      theme_minimal() +
      labs(x="Regression coefficient",y="Functional trait")

```

#### Predicted elements by urbanization

```{r element_urb_pred, results="hide", eval=TRUE, message=FALSE, warning=FALSE, cache=TRUE}

community_elem_urb <- pred_urban %>%
  group_by(index500, genome) %>%
  mutate(row_id = row_number()) %>%
  pivot_wider(names_from = genome, values_from = value) %>%
  ungroup() %>%
  group_split(row_id) %>%
  as.list() %>%
  lapply(., FUN = function(x){x %>%
    select(-row_id) %>%
    column_to_rownames(var = "index500") %>%
    as.data.frame() %>%
    exp() %>%
    t() %>%
    tss() %>%
    to.community(GIFTs_elements,.,GIFT_db) %>%
    as.data.frame() %>%
    rownames_to_column(var="index500")
   })


calculate_slope <- function(x) {
  lm_fit <- lm(unlist(x) ~ seq_along(unlist(x)))
  coef(lm_fit)[2]
}

element_pred_urb <- map_dfc(community_elem_urb, function(mat) {
      mat %>%
        column_to_rownames(var = "index500") %>%
        t() %>%
        as.data.frame() %>%
        rowwise() %>%
        mutate(slope = calculate_slope(c_across(everything()))) %>%
        select(slope) }) %>%
      t() %>%
      as.data.frame() %>%
      set_names(colnames(community_elem_urb[[1]])[-1]) %>%
      rownames_to_column(var="iteration") %>%
      pivot_longer(!iteration, names_to="trait",values_to="value") %>%
      group_by(trait) %>%
      summarise(mean=mean(value),
        p1 = quantile(value, probs = 0.1),
        p9 = quantile(value, probs = 0.9),
        positive_support = sum(value > 0)/1000,
        negative_support = sum(value < 0)/1000) %>%
      arrange(-positive_support)

```

```{r element_urb_tables, message=FALSE, warning=FALSE}

# Positively associated
element_pred_urb %>%
  filter(mean >0) %>%
  arrange(-positive_support) %>%
  filter(positive_support>=0.9) %>%
  paged_table()

#Negatively associated
element_pred_urb %>%
  filter(mean <0) %>%
  arrange(-negative_support) %>%
  filter(negative_support>=0.9) %>%
  paged_table()


```

```{r element_urb_plot, message=FALSE, warning=FALSE, fig.dim=c(10, 10)}
#Positively associated
positive <- element_pred_urb%>%
  filter(mean >0) %>%
  arrange(mean) %>%
  filter(positive_support>=0.9) %>%
  select(-negative_support) %>%
  rename(support=positive_support)

#Negatively associated
negative <- element_pred_urb %>%
  filter(mean <0) %>%
  arrange(mean) %>%
  filter(negative_support>=0.9) %>%
  select(-positive_support) %>%
  rename(support=negative_support)

all_elements <- bind_rows(positive,negative) %>%
  left_join(GIFT_db,by=join_by(trait==Code_element)) %>%
  mutate(trait=factor(trait,levels=c(rev(positive$trait),rev(negative$trait)))) %>%
  mutate(Code_function=factor(Code_function)) %>%
  mutate(element_legend=str_c(trait," - ",Element)) %>%
  mutate(function_legend=str_c(Code_function," - ",Function)) %>%
  select(trait,mean,p1,p9,element_legend,function_legend) %>% 
  unique()

gift_colors <- gift_colors %>% 
  mutate(legend=str_c(Code_function," - ",Function))  %>% 
  filter(legend %in% all_elements$function_legend)

all_elements %>%
  ggplot(aes(x=mean, y=fct_reorder(element_legend, mean), xmin=p1, xmax=p9, color=function_legend)) +
      geom_point() +
      geom_errorbar() +
      #xlim(c(-0.05,0.05)) +
      geom_vline(xintercept=0) +
      scale_color_manual(values = gift_colors$Color) +
      theme_minimal() +
      labs(x="Regression coefficient",y="Functional trait")

community_elem_urb %>%
    bind_rows() %>%
    pivot_longer(-index500, names_to = "trait", values_to = "value") %>%
    filter(trait %in% c(positive$trait, negative$trait)) %>%
    mutate(trait=factor(trait, levels=c(positive$trait, negative$trait))) %>%
    mutate(index500=as.numeric(index500)) %>%
    ggplot(aes(x=index500, y=value)) +
          geom_smooth(method = lm, formula = y ~ x, se = TRUE) +
          #geom_smooth(method = lm, formula = y ~ splines::bs(x, 3), se = TRUE) +
          facet_wrap(~trait, ncol=6, scales="free") +
          theme_minimal() +
          labs(x="Urbanization",y="Metabolic Capacity Index")
```

#### Predicted elements by season

**S. carolinensis**

**S. vulgaris**





