---
title: "AlberdiLab | Romeo et al. 2024"
subtitle: "Study title to be added"
author:
  - Claudia Romeo^[University of Copenhagen, claudia.romeo@sund.ku.dk], [...], and Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
date: "Last update: `r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/invasion_hologenomics
description: |
  Data analysis code for the study of metagenome‑assembled genomes and derived microbial communities from faecal samples of invasive and native squirrel species.
link-citations: yes
github-repo: alberdilab/invasion_hologenomics
always_allow_html: yes
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for the analysis of metagenomic data from invasive Sciurus carolinensis and native Sciurus vulgaris squirrels along urbanization gradients and across seasons.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/invasion_hologenomics.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}

# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
library(rmarkdown)

# For tree handling
library(ape)
library(phyloseq)
library(phytools)

# For plotting
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(ggnewscale)
library(gridExtra)
library(ggtreeExtra)
library(ggtree)
library(ggh4x)
library(mapproj)
library(RColorBrewer)
library(reshape2)
library(ggResidpanel)
library(tidybayes)

# For statistics
library(vegan)
library(Rtsne)
library(hilldiv2)
library(distillR)
library(broom.mixed)
library(gtsummary)
library(car)
library(Hmsc)
library(corrplot)
library(brms)
library(parameters)
library(bayestestR)
library(pairwiseAdonis) 

```

<!--chapter:end:index.Rmd-->


# Data preparation

Placeholder


#### Read counts
#### Genome coverage
#### Sample metadata
#### Genome metadata
#### Coverage filtering
#### Generate genome count table
#### Genome tree
#### MAG functional annotations
#### Colour palettes
#### Wrap working objects

<!--chapter:end:00-data_preparation.Rmd-->


# Data Summary

Placeholder


## Sample summary
## Sequencing data summary
## MAGs summary

<!--chapter:end:01-data_summary.Rmd-->


# MAGs overview

Placeholder


## MAGs phylogeny
## Genome quality
## Functional attributes of MAGs
## Functional ordination of MAGs (distillr)

<!--chapter:end:02-MAGs_overview.Rmd-->


# Community composition

Placeholder


## Count data preparation
## Genomes by host species
## Taxonomic composition of samples

<!--chapter:end:03-community_composition.Rmd-->


# Diversity analyses

Placeholder


## Data preparation
## Alpha Diversity 
### Alpha diversity comparisons
### Alpha diversity bayesian models
## Beta diversity
### Both species
### Red squirrel
### Grey squirrel

<!--chapter:end:04-diversity_analysis.Rmd-->


# Functional analyses

Placeholder


## Data preparation
## Genome-specific GIFT profiles
## Element-level community-averaged GIFT profiles 
## Function-level community-averaged GIFT profiles
## Domain-level community-averaged GIFT profiles

<!--chapter:end:05-functional_analysis.Rmd-->


# Joint Species Distribution Modelling - model setup

Placeholder


## Prepare input for Hmsc
## Define formulas of the Hmsc model
## Define and Hmsc models
## Define MCMC
## Generate Hmsc executables
## Fit Hmsc models (in Mjolnir HPC)
## Assess chain convergence

<!--chapter:end:06-hmsc_setup.Rmd-->


# Joint Species Distribution Modelling - output analysis

Placeholder


## Variance partitioning
## Posterior estimates
## Model fit
## Correlations among bacteria
## Association with host species
### MAGs associated with **S. carolinensis**
### MAGs associated with **S. vulgaris**
## Microbiome composition predictions
### Predicted composition by species
### Predicted composition by urbanization
### Predicted composition by season
## Microbiome functional predictions
### Function-level predictions
#### Predicted functions by species
#### Predicted functions by urbanization
#### Predicted elements by season
### Element-level predictions
#### Predicted elements by species
#### Predicted elements by urbanization
#### Predicted elements by season

<!--chapter:end:07-hmsc_analysis.Rmd-->

