---
title: "AlberdiLab | Romeo et al. 2024"
subtitle: "Study title to be added"
author:
  - Claudia Romeo^[University of Copenhagen, claudia.romeo@sund.ku.dk], [...], and Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
date: "Last update: `r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/invasion_hologenomics
description: |
  Data analysis code for the study of metagenome‑assembled genomes and derived microbial communities from faecal samples of invasive and native squirrel species.
link-citations: yes
github-repo: alberdilab/invasion_hologenomics

---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for the analysis of metagenomic data from invasive Sciurus carolinensis and native Sciurus vulgaris squirrels along urbanization gradients and across seasons.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/invasion_hologenomics.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}

# Base
library(devtools)
library(knitr)
library(rmarkdown)
library(R.utils)
library(tinytable)
library(tidyverse)
#library(easystats)

# For tree handling
library(ape)
library(phyloseq)
library(phytools)

# For plotting
library(aplot)
library(ggh4x)
library(ggnewscale)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(ggResidpanel)
library(ggtree)
library(ggtreeExtra)
library(ggvenn)
library(gridExtra)
library(mapproj)
library(RColorBrewer)
library(reshape2)
library(tidybayes)

# For analysis
library(ANCOMBC)
library(bayestestR)
library(brms)
library(broom.mixed)
library(car)
library(corrplot)
library(distillR)
library(geosphere)
library(gtsummary)
library(hilldiv2)
library(Hmsc)
library(iNEXT)
library(microbiome)
library(pairwiseAdonis) 
library(parameters)
library(Rtsne)
library(vegan)

```
